---
title: "Student Adaptability Level Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Student Adaptability Level Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/mdmahmudulhasansuzan/students-adaptability-level-in-online-education\>*

### Reference:

*\<Md. Mahmudul Hasan Suzan. (n.d.). Students’ Adaptability Level in Online Education [Data set]. Kaggle. https://www.kaggle.com/datasets/mdmahmudulhasansuzan/students-adaptability-level-in-online-education\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*


# Understanding the Dataset (Exploratory Data Analysis (EDA))

## Loading the Dataset
```{r Load dataset}
# Load the dataset
EducationData <- read.csv("students_adaptability_level_online_education.csv", colClasses = c(
  Gender = "factor",
  Education_Level = "factor",
  Institution_Type = "factor",
  IT_Student = "factor",
  Location = "factor",
  Load_shedding = "factor",
  Financial_Condition = "factor",
  Internet_Type = "factor",
  Network_Type = "factor",
  Self_Lms = "factor",
  Device = "factor",
  Adaptivity_Level = "factor"
))

# Display structure to verify data types
str(EducationData)

# Display first few rows to ensure data is loaded correctly
head(EducationData)

# View the dataset in a spreadsheet-like interface (optional)
View(EducationData)
```

## Measures of Frequency
```{r MOF}
# Load required libraries
library(DescTools)  # For Mode function

# 1. Measures of Frequency
gender_frequency <- table(EducationData$Gender)
education_level_frequency <- table(EducationData$Education_Level)

print("Frequency Tables:")
print(gender_frequency)
print(education_level_frequency)

# Relative frequency
gender_relative_freq <- prop.table(gender_frequency)
education_level_relative_freq <- prop.table(education_level_frequency)

print("Relative Frequencies:")
print(gender_relative_freq)
print(education_level_relative_freq)
```

## Measures of Central Tendency
```{r MOCT}
# 2. Measures of Central Tendency
# Recode Adaptivity_Level to numeric for numerical operations
EducationData$Adaptivity_Level <- as.numeric(factor(EducationData$Adaptivity_Level, 
                                                    levels = c("Low", "Moderate", "High")))

mean_adaptivity <- mean(EducationData$Adaptivity_Level, na.rm = TRUE)
median_adaptivity <- median(EducationData$Adaptivity_Level, na.rm = TRUE)
mode_adaptivity <- Mode(EducationData$Adaptivity_Level, na.rm = TRUE)

print(paste("Mean Adaptivity Level:", mean_adaptivity))
print(paste("Median Adaptivity Level:", median_adaptivity))
print(paste("Mode Adaptivity Level:", mode_adaptivity))
```

## Measures of Distribution
```{r MOD}
# 3. Measures of Distribution
range_adaptivity <- range(EducationData$Adaptivity_Level, na.rm = TRUE)
variance_adaptivity <- var(EducationData$Adaptivity_Level, na.rm = TRUE)
sd_adaptivity <- sd(EducationData$Adaptivity_Level, na.rm = TRUE)
quantiles_adaptivity <- quantile(EducationData$Adaptivity_Level, na.rm = TRUE)

print(paste("Range:", range_adaptivity[1], "to", range_adaptivity[2]))
print(paste("Variance:", variance_adaptivity))
print(paste("Standard Deviation:", sd_adaptivity))
print("Quantiles:")
print(quantiles_adaptivity)
```

## Measures of Relationship
```{r MOR}
# 4. Measures of Relationship
# Correlation (numerical relationship)
correlation <- cor(as.numeric(EducationData$Financial_Condition), 
                   EducationData$Adaptivity_Level, 
                   use = "complete.obs")

print(paste("Correlation between Financial Condition and Adaptivity Level:", correlation))

# Contingency table (categorical relationship)
contingency_table <- table(EducationData$Gender, EducationData$Device)
print("Contingency Table:")
print(contingency_table)

# Chi-squared test for independence
chi_sq_test <- chisq.test(contingency_table)
print("Chi-squared Test Result:")
print(chi_sq_test)

```

## Plots
```{r Plots}
# Load required libraries
library(ggplot2)

# 1. Univariate Plots
# Histogram for a numerical variable (Adaptivity_Level)
ggplot(EducationData, aes(x = Adaptivity_Level)) +
  geom_histogram(binwidth = 0.5, fill = "blue", color = "black") +
  labs(title = "Histogram of Adaptivity Level", x = "Adaptivity Level", y = "Count") +
  theme_minimal()

# Bar plot for a categorical variable (Gender)
ggplot(EducationData, aes(x = Gender)) +
  geom_bar(fill = "green", color = "black") +
  labs(title = "Bar Plot of Gender", x = "Gender", y = "Count") +
  theme_minimal()

# Boxplot for a numerical variable (Adaptivity_Level by Gender)
ggplot(EducationData, aes(x = Gender, y = Adaptivity_Level, fill = Gender)) +
  geom_boxplot() +
  labs(title = "Boxplot of Adaptivity Level by Gender", x = "Gender", y = "Adaptivity Level") +
  theme_minimal()

# 2. Multivariate Plots
# Scatter plot (Financial Condition vs. Adaptivity Level)
ggplot(EducationData, aes(x = as.numeric(Financial_Condition), y = Adaptivity_Level)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Scatter Plot of Financial Condition vs. Adaptivity Level", 
       x = "Financial Condition", y = "Adaptivity Level") +
  theme_minimal()

# Stacked bar plot (Gender and Device)
ggplot(EducationData, aes(x = Gender, fill = Device)) +
  geom_bar(position = "stack") +
  labs(title = "Stacked Bar Plot of Gender and Device", x = "Gender", y = "Count") +
  theme_minimal()

# Faceted plot (Adaptivity Level by Internet Type)
ggplot(EducationData, aes(x = Internet_Type, y = Adaptivity_Level, fill = Internet_Type)) +
  geom_boxplot() +
  facet_wrap(~ Gender) +
  labs(title = "Boxplot of Adaptivity Level by Internet Type (Faceted by Gender)",
       x = "Internet Type", y = "Adaptivity Level") +
  theme_minimal()


```

# Preprocessing and Data Transformation
## Missing Values
```{r missingness}
# Check for missing values in the dataset
missing_values_summary <- colSums(is.na(EducationData))

# Display the summary of missing values per column
print("Summary of Missing Values:")
print(missing_values_summary)

# Check if the dataset contains any missing values
any_missing <- any(is.na(EducationData))

if (any_missing) {
  print("The dataset contains missing values.")
} else {
  print("The dataset has no missing values.")
}

```

# Training Model
## Data Splitting
```{r Data Splitting}
# Load required libraries
library(caret)  # For cross-validation and data splitting
library(boot)   # For bootstrapping

# Load dataset
set.seed(123)  # Set seed for reproducibility

# 1. Data Splitting
# Split data into training (70%) and testing (30%) sets
train_index <- createDataPartition(EducationData$Adaptivity_Level, p = 0.7, list = FALSE)
train_data <- EducationData[train_index, ]
test_data <- EducationData[-train_index, ]

print("Training and Testing Data Summary:")
print(paste("Training Data Rows:", nrow(train_data)))
print(paste("Testing Data Rows:", nrow(test_data)))

```

## Bootstrapping
```{r Bootstrapping}
# 2. Bootstrapping
# Define a function to calculate a statistic (e.g., mean of Adaptivity_Level)
bootstrap_function <- function(data, indices) {
  d <- data[indices, ]  # Resample with replacement
  return(mean(d$Adaptivity_Level, na.rm = TRUE))
}

# Apply bootstrapping with 1000 resamples
bootstrap_results <- boot(data = train_data, statistic = bootstrap_function, R = 1000)

print("Bootstrapping Results:")
print(bootstrap_results)
```

## Cross-validation
```{r Cross-validation}
# 3. Cross-validation
# Basic k-fold cross-validation (k = 10)
cv_control <- trainControl(method = "cv", number = 10)
cv_model <- train(Adaptivity_Level ~ ., data = train_data, method = "lm", trControl = cv_control)

print("Basic Cross-validation Results:")
print(cv_model)

# Repeated k-fold cross-validation (k = 10, repeated 5 times)
repeated_cv_control <- trainControl(method = "repeatedcv", number = 10, repeats = 5)
repeated_cv_model <- train(Adaptivity_Level ~ ., data = train_data, method = "lm", trControl = repeated_cv_control)

print("Repeated Cross-validation Results:")
print(repeated_cv_model)

# Leave-One-Out Cross-Validation (LOOCV)
loocv_control <- trainControl(method = "LOOCV")
loocv_model <- train(Adaptivity_Level ~ ., data = train_data, method = "lm", trControl = loocv_control)

print("LOOCV Results:")
print(loocv_model)
```

## Training Different Models
```{r Training Different Models}
# Load required libraries
library(caret)       # For data splitting and model training
library(rpart)       # For Decision Tree
library(randomForest) # For Random Forest

# Set seed for reproducibility
set.seed(123)

# Split data into training and testing sets (70-30 split)
train_index <- createDataPartition(EducationData$Adaptivity_Level, p = 0.7, list = FALSE)
train_data <- EducationData[train_index, ]
test_data <- EducationData[-train_index, ]

# 1. Linear Regression
linear_model <- train(
  Adaptivity_Level ~ ., 
  data = train_data, 
  method = "lm",
  trControl = trainControl(method = "cv", number = 10)  # 10-fold CV
)

print("Linear Regression Model:")
print(linear_model)

# 2. Decision Tree Regression
tree_model <- train(
  Adaptivity_Level ~ ., 
  data = train_data, 
  method = "rpart",
  trControl = trainControl(method = "cv", number = 10)  # 10-fold CV
)

print("Decision Tree Regression Model:")
print(tree_model)

# 3. Random Forest Regression
rf_model <- train(
  Adaptivity_Level ~ ., 
  data = train_data, 
  method = "rf",
  trControl = trainControl(method = "cv", number = 10),  # 10-fold CV
  tuneLength = 5  # Test 5 different values of mtry
)

print("Random Forest Regression Model:")
print(rf_model)

# Evaluate all models on the test data
linear_predictions <- predict(linear_model, newdata = test_data)
tree_predictions <- predict(tree_model, newdata = test_data)
rf_predictions <- predict(rf_model, newdata = test_data)

# Calculate performance metrics for all models
linear_rmse <- RMSE(linear_predictions, test_data$Adaptivity_Level)
tree_rmse <- RMSE(tree_predictions, test_data$Adaptivity_Level)
rf_rmse <- RMSE(rf_predictions, test_data$Adaptivity_Level)

print(paste("Linear Regression RMSE:", linear_rmse))
print(paste("Decision Tree RMSE:", tree_rmse))
print(paste("Random Forest RMSE:", rf_rmse))
```

## Model Performance Comparison using resamples
```{r Performance Comparison}
# Compare model performance using resamples
models <- list(
  Linear_Regression = linear_model,
  Decision_Tree = tree_model,
  Random_Forest = rf_model
)

# Resample models to compare performance
resample_results <- resamples(models)

# Summary of resampling results
summary(resample_results)

# Boxplot for performance metrics
bwplot(resample_results, layout = c(3, 1))

# Dot plot for performance metrics
dotplot(resample_results, layout = c(3, 1))

# Pairwise comparison of model performance (optional)
model_diff <- diff(resample_results)
summary(model_diff)


```

## Saving Model
```{r Saving Model}
# Saving the model
saveRDS(rf_model, "./models/saved_rf_model.rds")

# Load the saved model
loaded_rf_model <- readRDS("./models/saved_rf_model.rds")

# Example data for predictions
new_data <- data.frame(
  Gender = "Boy", 
  Education_Level = "University", 
  Institution_Type = "Non Government", 
  IT_Student = "No", 
  Location = "Yes", 
  Load_shedding = "Low", 
  Financial_Condition = "Mid", 
  Internet_Type = "Wifi", 
  Network_Type = "4G", 
  Self_Lms = "No", 
  Device = "Tab", 
  Adaptivity_Level = "Moderate"
)

# Use the loaded model to make predictions
predictions_loaded_model <- predict(loaded_rf_model, newdata = new_data)

# Print predictions
print(predictions_loaded_model)

prediction_numeric <- predict(loaded_rf_model, newdata = new_data)

# Map numeric prediction to factor levels
prediction_factor <- ifelse(prediction_numeric <= 1.5, "Low", 
                            ifelse(prediction_numeric <= 2.5, "Moderate", "High"))

# Print the prediction as a factor
print(prediction_factor)

```


